---
title: "SpNeigh analysis of mouse brain Xenium data"
author:
  - name: Jinming Cheng
    affiliation:
    - &duk_nus Centre for Quantitative Medicine,
               Duke-NUS Medical School,
               Singapore, 169857, 
               Singapore
    # email: jinming.cheng@outlook.com
    

date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{SpNeigh analysis of mouse brain Xenium data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

output:
  BiocStyle::html_document:
    toc_float: false
bibliography-style: plain
---

<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


# Mouse Brain Xenium

This Rmarkdown file includes the SpNeigh analysis of mouse brain tiny Xenium dataset to reproduce all the related figures in the SpNeigh manuscript.

## Data preparation

Load packages
```{r, message=FALSE, warning=FALSE}
library(SpNeigh)
library(Seurat)
library(ggplot2)
```


Read in processed Seurat object of mouse brain Xenium data
```{r}
xenium.obj <- readRDS("data/Mouse_brain_tiny_xenium_object.rds")
dim(xenium.obj)
```


```{r,include=FALSE}
figure_path = "figures/vigette_mouse_brain_xenium/"
knitr::opts_chunk$set(
    eval = TRUE,
    collapse = TRUE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 72,
    fig.height = 4.2,
    fig.width = 5.6,
    fig.path = figure_path
)
```

Extract coordinates from Xenium object
```{r}
coords <- SpNeigh::ExtractCoords(xenium.obj)
rownames(coords) <- coords$cell
head(coords)
```

Backup clusters 
```{r}
coords$raw_cluster <- coords$cluster
```


Define function to sort factor levels by decreasing number
```{r}
FactorSorted = function(x, decreasing = TRUE){
  x = factor(x)
  ordered_levels = sort(table(x),decreasing = decreasing)
  x = factor(x, levels = names(ordered_levels) )
  x
}
```


Cluster-level cell type
```{r}
new_cell_types = c("0"="Meninges",
                   "1"="Cerebral_cortex",
                   "2"="White_matter",
                   "3"="Cerebral_cortex",
                   "4"="Cerebral_cortex",
                   "5"="Thalamus",
                   "6"="Cerebral_cortex",
                   "7"="Hippocampus",
                   "8"="Hippocampus",
                   "9"="Hippocampus",
                   "10"="White_matter",
                   "11"="Cerebral_cortex")
coords$cluster_level_cell_type <- factor(coords$raw_cluster,
                                         levels = names(new_cell_types),
                                         labels = new_cell_types)
coords$cluster_level_cell_type <- FactorSorted(coords$cluster_level_cell_type)
table(coords$cluster_level_cell_type)
```

Read in SingleR annotation (predicted by `SingleR()`, code not shown)
```{r}
SingleR_cell_type <-readRDS("data/Mouse_brain_SingleR_cell_type.rds")
```


The raw cell types from SingleR prediction were manually merged 
to reduce redundancy.
```{r}
coords$SingleR_cell_type_raw <- SingleR_cell_type$raw_cell_type
coords$SingleR_cell_type_merged <- SingleR_cell_type$new_cell_type
```

The raw SingleR annotation has 42 cell types
```{r}
nlevels(coords$SingleR_cell_type_raw)
```
The manually merged SingleR annotation has 10 cell types, 
which was used in the main figure 2
```{r}
nlevels(coords$SingleR_cell_type_merged)
```


Extract normalized expression data
```{r}
logNorm_expr <- Seurat::GetAssayData(xenium.obj)
```

```{r}
class(logNorm_expr)
```


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->



<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


## SpNeigh analysis


### Figure 1 

(For illustration purpose)  

Figure 1b-1. Detect boundaries.
```{r fig1b_1, fig.width=5.6, fig.height=4.2}
boundary_points_c2 <- GetBoundary(data = xenium.obj, one_cluster = 2, eps=120)
PlotBoundary(coords, boundary = boundary_points_c2,
             sub_plot = TRUE, one_cluster = 2) +
  Seurat::NoAxes() + Seurat::NoLegend()
```

Figure 1b-2. Detect boundaries (Rings).
```{r fig1b_2, fig.width=5.6, fig.height=4.2}
outer_boundary_c2 <- GetOuterBoundary(boundary = boundary_points_c2)
PlotBoundary(coords, boundary = boundary_points_c2, 
             sub_plot = TRUE, one_cluster = 2) +
  AddBoundary(outer_boundary_c2, color_boundary = "blue") +
  Seurat::NoAxes() + Seurat::NoLegend()
  
```
Figure 1c-1. Boundary weights
```{r fig1c_1, fig.width=5.6, fig.height=4.2}
cells_c2 <- subset(coords, cluster == "2")[,"cell"]
bon_weights_c2 <- ComputeBoundaryWeights(coords, 
                                         cell_ids = cells_c2,
                                         boundary = boundary_points_c2)
PlotWeights(coords, 
            weights = bon_weights_c2, 
            point_size = 0.8) + 
  Seurat::NoAxes() + Seurat::NoLegend() + 
  ggplot2::ggtitle("Boundary weights")
```

Figure 1c-2. Centroid weights.
```{r fig1c_2, fig.width=5.6, fig.height=4.2}
cells_c2 <- subset(coords, cluster == "2")[,"cell"]
cen_weights_c2 <- ComputeCentroidWeights(coords, 
                                         cell_ids = cells_c2)
PlotWeights(coords, 
            weights = cen_weights_c2, 
            point_size = 0.8) + 
  NoAxes() + NoLegend() +  ggtitle("Centroid weights")
```

Figure 1d-1. Proportion of cells in neighborhood rings
```{r fig1d_1, fig.width=5.6, fig.height=3}
ring_c2 <- GetRingRegion(boundary = boundary_points_c2,
                         outer_boundary = outer_boundary_c2)
cells_in_ring_c2 <- GetCellsInside(coords,boundary = ring_c2)
cell_stats_ring_c2 <- StatsCellsInside(cells_in_ring_c2)
PlotStatsPie(cell_stats_ring_c2, add_labels = FALSE) + 
  Seurat::NoLegend()
```

Figure 1d-2. Interaction matrix
```{r fig1d_2, fig.width=5.6, fig.height=4.2}
interact_mat_ring_c2 <- 
  ComputeSpatialInteractionMatrix(
    subset(coords, cell %in% cells_in_ring_c2$cell)
    )
PlotInteractionMatrix(interact_mat_ring_c2)
```

Figure 1e-1. RunLimmaDE
```{r fig1e_1, fig.width=4, fig.height=3}
# uses tab results from RunLimmaDE from Figure 2
tab <- readRDS("data/Mouse_brain_tab_c2_RunLimmaDE.rds")

df <- tab[,c("logFC","adj.P.Val")]
df$trend <- ifelse(df$logFC>0,"positive","negative")

ggplot(data = df,
       aes(x=logFC,y= -log10(adj.P.Val), color=trend)) +
  geom_point() + my_theme_ggplot() + NoLegend() +
  labs(x="log2FC", y = "-log10(adj.Pval)", title = "RunLimmaDE" ) +
  coord_flip() +
  theme(plot.title=element_text(size=13, face="bold",hjust=0.5)) +
  scale_color_manual(values = scales::hue_pal()(2)[c(2,1)] )
```


Figure 1e-2. RunSpatialDE
```{r fig1e_2, fig.width=4, fig.height=3}
tab_bon_c2 <- RunSpatialDE(exp_mat = logNorm_expr[,cells_c2], 
                           cell_ids = cells_c2, 
                           spatial_distance = bon_weights_c2)

df <- tab_bon_c2[1:10,]
df <- dplyr::arrange(df, desc(abs(Z1)))
df <- dplyr::mutate(df, gene = factor(gene, levels = gene[order(Z1)]))

ggplot(df, aes(x = gene, y = Z1, fill = trend)) +
  geom_col() +
  coord_flip() +
  labs(x = NULL, y = "Z1", fill="Trend", title = "RunSpatialDE" ) + 
  theme_minimal() + NoLegend() + 
  theme(plot.title=element_text(size=13, face="bold",hjust=0.5)) +
  scale_fill_manual(values = scales::hue_pal()(2)[c(2,1)] )
```

Figure 1e-3. ComputeSEI
```{r fig1e_3, fig.width=4, fig.height=3}
sei_bon_c2 <- ComputeSpatialEnrichmentIndex(logNorm_expr[,cells_c2], 
                                            weights = bon_weights_c2)
df <- sei_bon_c2[1:10,]
df <- dplyr::mutate(df, 
                    gene = factor(gene, levels = df$gene) )

ggplot(df, aes(x = gene, y = (normalized_SEI - 1) )) +
  geom_col( fill = "mediumturquoise") +
  coord_flip() +
  labs(x = NULL, y = "Norm_SEI - 1", title = "ComputeSEI" ) + 
  theme_minimal() + NoLegend() + 
  theme(plot.title=element_text(size=13, face="bold",hjust=0.5))
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Figure 2

Figure 2a-1. Clusters.
```{r fig2a_1, fig.width=5.6, fig.height=3.5}
coords$cluster <- coords$raw_cluster
PlotBoundary(coords, boundary = boundary_points_c2) + NoAxes()
```

Figure 2a-2. Cluster-level annotation.
```{r fig2a_2, fig.width=6, fig.height=3.5}
# set cluster as cluster_level_cell_type
coords$cluster <- coords$cluster_level_cell_type
PlotBoundary(coords) + NoAxes()
```

Figure 2a-3. Cell-level annotation (manually merged).
```{r fig2a_3, fig.width=6, fig.height=3.5}
# set cluster as SingleR_cell_type_merged
coords$cluster <- coords$SingleR_cell_type_merged
PlotBoundary(coords) + NoAxes()
```

Set raw cluster back
```{r}
coords$cluster <- coords$raw_cluster
```


Figure 2b-1. Boundary regions.
```{r fig2b_1, fig.width=5.6, fig.height=2.8}
boundary_polys_c2 <- BuildBoundaryPoly(boundary_points_c2)
PlotRegion(boundary_poly = boundary_polys_c2) + NoAxes()
```

Figure 2b-2. Ring regions.
```{r fig2b_2, fig.width=5.6, fig.height=2.8}
PlotRegion(boundary_poly = ring_c2) + NoAxes()
```

Figure 2b-3. Cells inside boundaries.
```{r fig2b_3, fig.width=4.8, fig.height=2.8}
cells_in_bon_c2 <- GetCellsInside(coords, boundary = boundary_polys_c2)
PlotCellsInside(cells_in_bon_c2) + NoAxes() + NoLegend()
```

Figure 2b-4. Donut plot of stats of cells inside boundary.
```{r fig2b_4, fig.width=2, fig.height=2, out.width='50%'}
cells_in_bon_c2 <- GetCellsInside(coords, boundary = boundary_polys_c2)

cell_stats_bon_c2 <- StatsCellsInside(cells_in_bon_c2)
PlotStatsPie(subset(cell_stats_bon_c2, region_id=="1"), 
             plot_donut = TRUE,label_cutoff = 0.1) + 
  NoAxes() + NoLegend()
```

Figure 2b-5. Cells inside boundaries.
```{r fig2b_5, fig.width=4.8, fig.height=2.8}
PlotCellsInside(cells_in_ring_c2) + NoAxes() + NoLegend()
```

Figure 2b-6. Donut plot of stats of cells inside ring
```{r fig2b_6, fig.width=2, fig.height=2, out.width='50%'}
PlotStatsPie(subset(cell_stats_ring_c2, region_id=="1"), 
             plot_donut = TRUE,label_cutoff = 0.05) + 
  NoAxes() + NoLegend()
```

Figure 2c-1. Expression plot of Slc17a7
```{r fig2c_1, fig.width=4, fig.height=2.8}
# Obtain cluster 2 cells inside and outside boundary of sub region
cells_in <- subset(cells_in_bon_c2, region_id==1 & cluster==2)[["cell"]]
cells_out <- subset(cells_in_ring_c2, region_id==1 & cluster==2)[["cell"]]
tab <- RunLimmaDE(exp_mat = logNorm_expr,
                  min.pct = 0.25,
                  cells_reference = cells_in,
                  cells_target = cells_out)
tab[1:5,c("logFC", "adj.P.Val", "pct.reference", "pct.target")]

PlotExpression(data = coords, 
               exp_mat = logNorm_expr, 
               genes = "Slc17a7",
               sub_plot = TRUE,
               sub_cells = c(cells_in, cells_out) ) + 
  AddBoundary(boundary_polys_c2[1,], linewidth_boundary = 0.6) +
  NoAxes() + theme(legend.position = "bottom")
```

<!-- # save tab into rds file to reproduce figure 1e-1 -->
<!-- ```{r, include=FALSE} -->
<!-- saveRDS(tab, file = "data/Mouse_brain_tab_c2_RunLimmaDE.rds") -->
<!-- ``` -->


Figure 2c-2. Expression plot of Sox10.
```{r fig2c_2, fig.width=4, fig.height=2.8}
PlotExpression(data = coords, 
               exp_mat = logNorm_expr, 
               genes = "Sox10",
               sub_plot = TRUE,
               sub_cells = c(cells_in, cells_out) ) + 
  AddBoundary(boundary_polys_c2[1,], linewidth_boundary = 0.6) +
  NoAxes() + theme(legend.position = "bottom")
```

Figure 2d. Split boundary 1 into edges.
```{r fig2d, fig.width=5.6, fig.height=2.8}
boundary_edges_c2 <- SplitBoundaryPolyByAnchor(boundary_polys_c2[1,])
PlotEdge(boundary_edges_c2) + NoAxes() + labs(color=NULL)
```


Figure 2e. Boundary weights to edge 2 for cerebral cortex cells.
```{r fig2e, fig.width=5.6, fig.height=3.6}
cells_cortex <- subset(
  coords, cluster_level_cell_type=="Cerebral_cortex")$cell
bon_weights_edge2 <- ComputeBoundaryWeights(
  coords, 
  cell_ids = cells_cortex, 
  boundary = boundary_edges_c2[2,])

PlotWeights(coords, weights = bon_weights_edge2) + 
  NoAxes() +
  AddBoundary(boundary_edges_c2[2,], linewidth_boundary = 0.8)
```

Figure 2f. Plot expression along spatial weights.
```{r fig2f, fig.width=5.6, fig.height=3.6}
tab_sp <- RunSpatialDE(exp_mat = logNorm_expr,
                       cell_ids = cells_cortex,
                       spatial_distance = bon_weights_edge2)

PlotSpatialExpression(exp_mat = logNorm_expr[,cells_cortex],
                      genes = tab_sp$gene[1:10], 
                      spatial_distance = bon_weights_edge2)
```

Figure 2g. Expression of Ccn2 and Cplx3 in cell types near edge2.
```{r fig2g, fig.width=8, fig.height=4}
keep_celltypes = c("L4/5/6", "L6 Car3", "L6b",
                   "Astrocytes", "Oligodendrocytes")
cells_keep = subset(coords,SingleR_cell_type_merged %in% keep_celltypes)$cell

p1 <- PlotExpression(data = coords, 
                     exp_mat = logNorm_expr,
                     genes = c("Ccn2"),
                     sub_plot = TRUE, 
                     sub_cells = cells_keep, 
                     split_by = "SingleR_cell_type_merged",
                     ncol = 4) + NoAxes()

p2 <- PlotExpression(data = coords, 
                     exp_mat = logNorm_expr,
                     genes = c("Cplx3"),
                     sub_plot = TRUE, 
                     sub_cells = cells_keep, 
                     split_by = "SingleR_cell_type_merged",
                     ncol = 4) + NoAxes()

patchwork::wrap_plots(p1, p2, ncol = 1)
```



<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Figure 3

Figure 3a. Benchmarking runtime.
```{r record_runtime}
running_time <- list()

# Running time of RunLimmaDE
one_method <- "RunLimmaDE"
t_start <- Sys.time()
tab_list <- list()
my_clusters <- levels(coords$cluster_level_cell_type)
for (i in my_clusters) {
  cells_i <- coords$cell[coords$cluster_level_cell_type == i]
  cells_other <- coords$cell[coords$cluster_level_cell_type != i]
  
  tab_list[[i]] <- RunLimmaDE(exp_mat = logNorm_expr, 
                             cells_reference = cells_other,
                             cells_target = cells_i)
  tab_list[[i]]$gene <- rownames(tab_list[[i]]) 
  tab_list[[i]]$cluster = i 
}

t_end <- Sys.time()
t_diff <- difftime(t_end, t_start,units = "secs")
running_time[[one_method]][["t_start"]]  <- t_start
running_time[[one_method]][["t_end"]]  <- t_end
running_time[[one_method]][["t_diff"]]  <- t_diff

# Running time of find all markers
one_method = "FindAllMarkers"
xenium.obj$cluster_level_cell_type <- coords$cluster_level_cell_type
xenium.obj@active.ident <- xenium.obj$cluster_level_cell_type
t_start <- Sys.time()
markers <- FindAllMarkers(xenium.obj, min.pct = 0)
t_end <- Sys.time()
t_diff <- difftime(t_end, t_start,units = "secs")
running_time[[one_method]][["t_start"]]  <- t_start
running_time[[one_method]][["t_end"]]  <- t_end
running_time[[one_method]][["t_diff"]]  <- t_diff
```

```{r fig3a, fig.width=3, fig.height=3}
df <- do.call(rbind,lapply(running_time,as.data.frame))
a <- round(as.numeric(df$t_diff),2)
a <- cbind.data.frame("Method" = rownames(df), "RunningTime" = a)
a$Method <- factor(a$Method,levels = c("RunLimmaDE","FindAllMarkers"))
a$Package <-  factor(a$Method,
                     levels = c("RunLimmaDE","FindAllMarkers"), 
                     labels = c("SpNeigh","Seurat"))

ggplot(data=a, aes(x=Method, y=RunningTime,fill=Package)) + 
  geom_col() + 
  labs(x=NULL, y="Seconds", fill="") + 
  my_theme_ggplot() + 
  RotatedAxis() +
  geom_text(aes(label = RunningTime), vjust=-0.5, size=4) +
  ylim(c(0,35)) + 
  labs(title = "Running time")
```

Fiure 3b. Dot plot of top marker genes from RunLimmaDE
```{r fig3b, fig.width=8,fig.height=4}
top_tab <- lapply(tab_list, function(x){head(subset(x,logFC>0), n = 3 )})
top_tab <- do.call(rbind,top_tab)

my_genes <- unique(top_tab$gene)

df <- t(logNorm_expr[my_genes,])
df <- as.data.frame(df)
df$cluster <- coords$cluster_level_cell_type

library(tidyr)
library(dplyr)
# Pivot longer to tidy format
df_long <- df %>% 
  pivot_longer(cols = -cluster, names_to = "gene", values_to = "expr")

# Group by cluster and gene, then summarize
summary_df <- df_long %>%
  group_by(cluster, gene) %>%
  summarise(
    avg_expr = mean(expr),
    pct_expressed = mean(expr > 0) * 100,
    .groups = "drop"
  )

summary_df$gene = factor(summary_df$gene, levels = my_genes)

ggplot( data = summary_df, 
        aes(x=cluster,y=gene, size=pct_expressed, color=avg_expr) ) +
  geom_point() +
  my_theme_ggplot() +
   ggplot2::scale_color_viridis_c(
    option = "plasma",
    oob = scales::squish #  replaces out of bounds values with the nearest limit.
    ) + 
  scale_size( limits = c(0,100),  range = c(1,5) )+
  ggplot2::theme(axis.text.x = ggplot2::element_text(angle = 45, hjust = 1)) +
  labs(color = "AvgExp",size="Percent",x=NULL,y=NULL)  + coord_flip()

```

Figure 3c. Spearman correlation of two DE methods.
```{r fig3c, fig.width=8,fig.height=2}
markers_list <- split(markers,markers$cluster)

# prepare data for plot
df <- list()
my_clusters <- levels(coords$cluster_level_cell_type)
for (i in my_clusters) {
  # RunLimmaDE markers
  a <- tab_list[[i]]
  rownames(a) <- a$gene
  
  # FindAllMarkers markers
  b <- markers_list[[i]]
  rownames(b) <- b$gene
  
  common_genes <- intersect(a$gene,b$gene)
  # logFC
  df[[i]] <- 
    data.frame("RunLimmaDE" = a[common_genes,"logFC"],
               "FindAllMarkers" = b[common_genes,"avg_log2FC"],
               "Cluster" = i)
}

df = do.call(rbind,df)

# Plot correlation
coef_spearman = cor(df$RunLimmaDE,df$FindAllMarkers, method = "spearman")
coef_spearman = round(coef_spearman,2)

p_title <- bquote("Overall log2FC correlation: Spearman " 
                  * rho == .(coef_spearman))
ggplot(data = df, 
       aes(x = RunLimmaDE, y = FindAllMarkers, colour = Cluster)) +
  geom_point(size = 0.3) + facet_wrap(.~Cluster, ncol=5) + 
  geom_smooth(method = "lm",size=0.8, se = FALSE) +
  labs(title = p_title,
       x = "RunLimmaDE",
       y = "FindAllMarkers") +
  scale_color_manual(values = my_colors_15) +
  my_theme_ggplot() + 
  theme(axis.text=element_text(size=12),
        axis.title=element_text(size=13, face="bold"),
        plot.title=element_text(size=15, face="bold",hjust=0.5) ) +
  NoLegend() + theme(plot.margin = margin(t = 0.2, b = 0, r =  0.2, l = 0.2, "cm"))
```

Figure 3d-1. Boundary weights of C0 cells.
```{r fig3d_1, fig.width=5.6, fig.height=3.6}
cells_c0 <- subset(coords, cluster == "0")[,"cell"]
bon_points_c0 <- GetBoundary(data = coords, one_cluster = 0)
bon_weights_c0 <- ComputeBoundaryWeights(coords, cell_ids = cells_c0,
                                         boundary = bon_points_c0)
PlotWeights(coords, weights = bon_weights_c0) + NoAxes()
```

Figure 3d-2. Centroid weights of C0 cells.
```{r fig3d_2, fig.width=5.6, fig.height=3.6}
cen_weights_c0 <- ComputeCentroidWeights(coords, cell_ids = cells_c0)
PlotWeights(coords, weights = cen_weights_c0) + NoAxes()
```

Figure 3e-1. Spearman correlation of two spatial methods (boundary weights).
```{r fig3e_1, fig.width=4, fig.height=3}
tab_sp_c0 <- RunSpatialDE(exp_mat = logNorm_expr, cell_ids = cells_c0,
                          spatial_distance = bon_weights_c0)
sei_bon_c0 <- ComputeSpatialEnrichmentIndex(logNorm_expr[,cells_c0],
                                            weights = bon_weights_c0)
my_tab <- tab_sp_c0
my_sei <- sei_bon_c0

rownames(my_sei) <- my_sei$gene
common_genes <- intersect(my_sei$gene, my_tab$gene)

df <- cbind.data.frame(my_sei[common_genes,],
                       my_tab[common_genes,"Z1",drop = FALSE])

df$rank_normalized_SEI <- rank(-df$normalized_SEI)
df$rank_Z1 <- rank(-df$Z1)

df$rank_normalized_SEI <- rank(-df$normalized_SEI)
df$rank_Z1 <- rank(-df$Z1)

coef_spearman <- cor(df$rank_normalized_SEI, df$rank_Z1,method = "spearman")
coef_spearman <- round(coef_spearman, 2)

ggplot(data = df, 
       aes(x=rank_normalized_SEI,y=rank_Z1)) + 
    geom_point(size = 0.5) + 
    my_theme_ggplot() + 
    labs(title = bquote("Spearman " * rho == .(coef_spearman)),
         x = "Rank of Norm SEI",
         y = "Rank of Z1 (Boun)") + 
  theme(axis.title=element_text(size=12, face="bold")) +
  geom_smooth(method = "lm",se = FALSE)
```

Figure 3e-2. Spearman correlation of two spatial methods (centroid weights).
```{r fig3e_2, fig.width=4, fig.height=3}
tab_sp_cen_c0 <- RunSpatialDE(exp_mat = logNorm_expr, cell_ids = cells_c0,
                              spatial_distance = cen_weights_c0)
sei_cen_c0 <- ComputeSpatialEnrichmentIndex(logNorm_expr[,cells_c0],
                                            weights = cen_weights_c0)
my_tab <- tab_sp_cen_c0
my_sei <- sei_cen_c0

rownames(my_sei) <- my_sei$gene
common_genes <- intersect(my_sei$gene, my_tab$gene)

df <- cbind.data.frame(my_sei[common_genes,],
                       my_tab[common_genes,"Z1",drop = FALSE])

df$rank_normalized_SEI <- rank(-df$normalized_SEI)
df$rank_Z1 <- rank(-df$Z1)

coef_spearman <- cor(df$rank_normalized_SEI, df$rank_Z1,method = "spearman")
coef_spearman <- round(coef_spearman, 2)

ggplot(data = df, 
       aes(x=rank_normalized_SEI,y=rank_Z1)) + 
    geom_point(size = 0.5) + 
    my_theme_ggplot() + 
    labs(title = bquote("Spearman " * rho == .(coef_spearman)),
         x = "Rank of Norm SEI",
         y = "Rank of Z1 (Cen)") + 
  theme(axis.title=element_text(size=12, face="bold")) +
  geom_smooth(method = "lm",se = FALSE)
```
Figure 3f. Top spatial DE genes (boundary).
```{r fig3f, fig.width=5, fig.height=3}
df <- tab_sp_c0[1:10,]
df <- dplyr::arrange(df, desc(abs(Z1)))
df <- dplyr::mutate(df, gene = factor(gene, levels = gene[order(Z1)]))

ggplot(df, aes(x = gene, y = Z1, fill = trend)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Positive" = "firebrick1", 
                               "Negative" = "cornflowerblue")) +
  labs(x = NULL, y = "Z1", fill="Trend", 
       title = "Top spatial DE genes") +
  theme_minimal() +
  theme(plot.title=element_text(face="bold",  hjust=0.5)) 
```

Figure 3g. Top enriched genes from SEI (boundary)
```{r fig3g, fig.width=5, fig.height=3}
df <- sei_cen_c0[1:10,]
df <- dplyr::mutate(df, gene = factor(gene, levels = df$gene) )

ggplot(df, aes(x = gene, y = (normalized_SEI - 1) )) +
  geom_col( fill = "mediumturquoise") +
  coord_flip() +
  labs(x = NULL, y = "Norm_SEI - 1",
       title = "Top enriched genes (SEI)") +
  theme_minimal() +
  theme(plot.title=element_text(face="bold",  hjust=0.5)) 
```
<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Supplementary Figure 1

Figure S1a. Cell-level annotation (raw).
```{r figS1a, fig.width=8, fig.height=3.5}
ggsci_d3_20 <- ggsci::pal_d3("category20")(20)
my_colors_20 <- c(my_colors_15, 
                  "gray60","gray40","royalblue2","mediumblue","black")

# set cluster as SingleR_cell_type_raw
coords$cluster <- coords$SingleR_cell_type_raw
PlotBoundary(coords, colors = c(my_colors_20, ggsci_d3_20)) + 
  labs(color=NULL) +
  NoAxes()
```

Figure S1b. Expression of Slc17a7 and Sox10.
```{r figS1b, fig.width=8, fig.height=2.5}
my_genes = c("Slc17a7", "Sox10")

p <- list()
for (i in my_genes) {
  p[[i]] <- PlotExpression(data = coords, exp_mat = logNorm_expr,
                           genes = i) + 
    NoAxes()
}
patchwork::wrap_plots(p, ncol = 2)
```


Figure S1c. Expression of Ccn2 and Cplx3
```{r figS1c, fig.width=8, fig.height=2.5}
my_genes = c("Ccn2", "Cplx3")

p <- list()
for (i in my_genes) {
  p[[i]] <- PlotExpression(data = coords, exp_mat = logNorm_expr,
                           genes = i) + 
    NoAxes()
}
patchwork::wrap_plots(p, ncol = 2)
```
<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

## Session information

The information of current R session and packages used for this analysis is shown below.

```{r}
sessionInfo()
```


\pagebreak
