---
title: "SpNeigh analysis of human HER2+ breast cancer Xenium data"
author:
  - name: Jinming Cheng
    affiliation:
    - &duk_nus Centre for Quantitative Medicine,
               Duke-NUS Medical School,
               Singapore, 169857, 
               Singapore
    # email: jinming.cheng@outlook.com
    

date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{SpNeigh analysis of mouse brain Xenium data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

output:
  BiocStyle::html_document:
    toc_float: false
bibliography-style: plain
---


```{r,include=FALSE}
figure_path = "figures/vigette_human_breast_cancer_xenium/"
knitr::opts_chunk$set(
    eval = TRUE,
    collapse = TRUE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 72,
    fig.height = 4.2,
    fig.width = 5.6,
    fig.path = figure_path
)
```


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


# Human breast cancer Xenium

This Rmarkdown file includes the SpNeigh analysis of human HER2+ breast cancer Xenium dataset (Sample 1) to reproduce all the related figures in the SpNeigh manuscript.

## Data preparation

Load packages
```{r, message=FALSE, warning=FALSE}
library(SpNeigh)
library(Seurat)
library(ggplot2)
library(dplyr)
```


Read in processed Seurat object of human breast cancer Xenium data 
from [Cheng et al](https://doi.org/10.1186/s12859-025-06044-0)
```{r}
xenium.obj <- readRDS("data/Human_breast_cancer_sample1_xenium_object.rds")
dim(xenium.obj)
```

The SingleR annotation is used for SpNeigh manuscript, and it is
available at: https://github.com/jinming-cheng/cheng_annotation_bmc_bioinfo.
```{r}
xenium.obj$seurat_clusters <- xenium.obj$predicted_cell_type_SingleR
```


Extract coordinates from Xenium object
```{r}
coords <- SpNeigh::ExtractCoords(xenium.obj)
rownames(coords) <- coords$cell
head(coords)
```

Define function to sort factor levels by decreasing number
```{r}
FactorSorted = function(x, decreasing = TRUE){
  x = factor(x)
  ordered_levels = sort(table(x),decreasing = decreasing)
  x = factor(x, levels = names(ordered_levels) )
  x
}
```

Relevel cluster by decreasing number
```{r}
coords$cluster <- FactorSorted(coords$cluster)
table(coords$cluster)
```

Set colors
```{r}
my_colors <- c("#6495ED", "#FF3030", "#FFD700", "#BF3EFF", "#00FA9A", 
               "#48D1CC", "#ADFF2F", "#FFA500", "gray40",  "#8B6914",
               "#FFC0CB", "gray60",  "#FF00FF", "#EE82EE", "#00FFFF",
               "#E5E5E5")
names(my_colors) <- levels(coords$cluster)
```


Extract normalized expression data
```{r}
logNorm_expr <- Seurat::GetAssayData(xenium.obj)
```

```{r}
class(logNorm_expr)
```


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->



<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


## SpNeigh analysis


### Figure 4 

Figure 4a-1. Spatial plot with neighborhood rings.
```{r fig4a_1, fig.width=6, fig.height=2.8}
bon_points_tumor <- GetBoundary(coords, one_cluster = "Tumor")
bon_polys_tumor <- BuildBoundaryPoly(bon_points_tumor)
rings_tumor <- GetRingRegion(bon_polys_tumor)

bon_points_dcis <- GetBoundary(coords, one_cluster = "DCIS")
bon_polys_dcis <- BuildBoundaryPoly(bon_points_dcis)
rings_dcis <- GetRingRegion(bon_polys_dcis)


PlotBoundary(coords, colors = my_colors) + 
  AddBoundaryPoly(rings_tumor, 
                  linewidth_boundary = 0.6, 
                  color_boundary = "blue") +
  AddBoundaryPoly(rings_dcis, 
                  linewidth_boundary = 0.6, 
                  color_boundary = "yellow2" ) +
  guides(color=guide_legend( ncol=2, override.aes = list(size=2) ) ) +
  Seurat::NoAxes()
```



Figure 4a-2. Neighborhood rings.
```{r fig4a_2, fig.width=4, fig.height=2.8}
ggplot() + 
  AddBoundaryPoly(rings_tumor, 
                  linewidth_boundary = 0.4, 
                  color_boundary = "blue") +
  AddBoundaryPoly(rings_dcis, 
                  linewidth_boundary = 0.4, 
                  color_boundary = "yellow2" ) +
  theme_minimal() +
  Seurat::NoAxes()
```


Obtain cells insides tumor rings (takes about 5 minutes)
```{r}
t_start <- Sys.time()
cells_inside_ring_tumor <- GetCellsInside(coords, boundary = rings_tumor)
t_end <- Sys.time()
t_diff <- difftime(t_end, t_start,units = "mins")
round(t_diff,2) 
```

Figure 4b-1. Plot cells in neighborhood rings (Tumor).
```{r fig4b_1, fig.width=4, fig.height=2.8}
PlotCellsInside(cells_inside_ring_tumor, point_size = 0.1) + 
  NoAxes() + NoLegend()
```

Obtain cells insides dcis rings (takes about 0.8 minutes)
```{r}
t_start <- Sys.time()
cells_inside_ring_dcis <- GetCellsInside(coords, boundary = rings_dcis)
t_end <- Sys.time()
t_diff <- difftime(t_end, t_start,units = "mins")
round(t_diff,2) 
```

Figure 4b-2. Plot cells inside neighborhood rings (DCIS).
```{r fig4b_2, fig.width=4, fig.height=2.8}
PlotCellsInside(cells_inside_ring_dcis, point_size = 0.1) + 
  NoAxes() + NoLegend()
```


Figure 4c-1. Cell type proportion in neighborhood (Tumor).
```{r fig4c_1, fig.width=3, fig.height=2.8, out.width="60%"}
cell_stats_rings_tumor <- StatsCellsInside(cells_inside_ring_tumor)
cell_stats_rings_tumor_combined <- cell_stats_rings_tumor  %>%
  group_by(cluster) %>%
  summarise(total = sum(count)) %>% 
  mutate(proportion = total/sum(total), 
         region_id = "combined") %>% 
  rename("count" = "total")

PlotStatsPie(cell_stats_rings_tumor_combined, plot_donut = TRUE, 
             label_nudge_x = 0.05, colors = my_colors, 
             label_cutoff = 0.02) + 
  theme(strip.text = element_blank()) + NoLegend()
```

Figure 4c-2. Cell type proportion in neighborhood (DCIS).
```{r fig4c_2, fig.width=3, fig.height=2.8, out.width="60%"}
cell_stats_rings_dcis <- StatsCellsInside(cells_inside_ring_dcis)
cell_stats_rings_dcis_combined <- cell_stats_rings_dcis  %>%
  group_by(cluster) %>%
  summarise(total = sum(count)) %>% 
  mutate(proportion = total/sum(total), 
         region_id = "combined") %>% 
  rename("count" = "total")

PlotStatsPie(cell_stats_rings_dcis_combined, plot_donut = TRUE, 
             label_nudge_x = 0.05, colors = my_colors, 
             label_cutoff = 0.02) + 
  theme(strip.text = element_blank()) + NoLegend()
```

Figure 4d-1. Cell type interaction in neighborhood (Tumor).
```{r fig4d_1, fig.width=5.6, fig.height=4.2}
interaction_mat_rings_tumor <- ComputeSpatialInteractionMatrix(
  subset(coords, cell %in% cells_inside_ring_tumor$cell))

PlotInteractionMatrix(interaction_mat_rings_tumor) +
  labs(title = "Tumor Neighborhood")
```

Figure 4d-2. Cell type interaction in neighborhood (DCIS).
```{r fig4d_2, fig.width=5.6, fig.height=4.2}
interaction_mat_rings_dcis <- ComputeSpatialInteractionMatrix(
  subset(coords, cell %in% cells_inside_ring_dcis$cell))

PlotInteractionMatrix(interaction_mat_rings_dcis) +
  labs(title = "Tumor Neighborhood")
```


Figure 4e. Spatial distribution of major immune cell types.
```{r fig4e, fig.width=8, fig.height=2}
df <- coords[coords$cluster %in% c("Macrophage", "T","B","Plasma"),]
PlotBoundary(df, point_size = 0.01,
             split_by = "cluster", ncol = 4,
             colors = my_colors) + 
  AddBoundaryPoly(bon_polys_tumor, color_boundary = "blue",  
                  linewidth_boundary = 0.6) +
  AddBoundaryPoly(bon_polys_dcis, color_boundary = "yellow2",  
                  linewidth_boundary = 0.6) + 
  NoAxes() +
  NoLegend()
```

Differential proportion tests between tumor and DCIS neighborhood using t.test, 
p.value for T cells is `0.012`
```{r}
res_test_prop <- list()
for (i in c(levels(coords$cluster))) {
  test_prop <- t.test(subset(cell_stats_rings_tumor, cluster==i)$proportion,
                      subset(cell_stats_rings_dcis, cluster==i)$proportion)
  res_test_prop[[i]] <- c(cluster=i, p_value = round(test_prop$p.value,3))
}

res_test_prop = as.data.frame(do.call(rbind, res_test_prop))
res_test_prop = res_test_prop[order(res_test_prop$p_value),]
res_test_prop
```


Figure 4f. Bar plot comparing T cell proportions between 
tumor and DCIS neighbor rings.
```{r fig4f, fig.width=8, fig.height=2}
# tumor neighborhood T cells
df1 <- subset(cell_stats_rings_tumor, cluster=="T")[,c("region_id", "proportion")]
df1$Rings <- "Tumor"
df1$region_id <- paste0("r1_",df1$region_id)

# DCIS neighborhood T cells
df2 <- subset(cell_stats_rings_dcis, cluster=="T")[,c("region_id", "proportion")]
df2$Rings <- "DCIS"
df2$region_id <- paste0("r2_",df2$region_id)

df <- rbind(df1, df2)
df$Rings = factor(df$Rings, levels = c("Tumor", "DCIS"))
df$region_id = FactorNaturalOrder(df$region_id)

ggplot(data = df, 
       aes(x = region_id, y = proportion, fill = Rings) ) + 
  geom_col() + my_theme_ggplot() + 
  theme( axis.ticks.x  = element_blank(),
         axis.text.x = element_blank() ) +
  labs(x = NULL, y = "Proportion",
       title = bquote("T cells (p =" ~ .("0.012") * ")") ) +
  scale_fill_manual(values = c("blue", "yellow2")) + 
  ggplot2::scale_y_continuous(labels = scales::percent_format())
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Figure 5

Figure 5a. Spatial plot of tumor and DCIS cells.
```{r fig5a, fig.width=4, fig.height=2.8}
PlotBoundary(coords[subset(coords, cluster %in% c("Tumor","DCIS"))$cell,],
             point_size = 0.1, 
             colors = my_colors) + 
  NoAxes() + NoLegend() +
  AddBoundary(bon_polys_tumor, color_boundary = "blue",  
              linewidth_boundary = 1) +
  AddBoundary(bon_polys_dcis, color_boundary = "yellow2",
              linewidth_boundary = 1)
```

Figure 5b. DCIS location relative to tumor.
```{r fig5b, fig.width=4, fig.height=2.8}
new_polys <- bon_polys_dcis
new_polys$region_id <- ifelse(new_polys$region_id==1,"In","Out")

PlotRegion(new_polys) + NoAxes()
```

DE analysis between DCIS cells inside and outside tumor
```{r}
cells_dcis_in_tumor <- GetCellsInside(data = coords, 
                                      boundary = bon_polys_dcis[1,])

cells_dcis_out_tumor <- GetCellsInside(data = coords, 
                                       boundary = bon_polys_dcis[-1,])

cells_ref <- cells_dcis_out_tumor$cell
cells_tar <- cells_dcis_in_tumor$cell
tab_de <- RunLimmaDE(exp_mat = logNorm_expr[,c(cells_ref, cells_tar)],
                     cells_reference = cells_ref,
                     cells_target = cells_tar)
head(tab_de)[,c(1,5,7,8)]
```


Figure 5c. Expression of TENT5C and MZB1 in tumor and DCIS cells.
```{r fig5c, fig.width=6.4, fig.height=4.8}
cells_tumor <- subset(coords, cluster=="Tumor")$cell
cells_DICS <- subset(coords, cluster=="DCIS")$cell

my_genes = c("TENT5C", "MZB1")
p <- list()
for (i in my_genes) {

  p[[i]] <- 
    PlotExpression(data = coords,
                   exp_mat = logNorm_expr,
                   genes = i,
                   split_by = "cluster", 
                   point_size = 0.1,
                   sub_plot = TRUE,
                   sub_cells = c(cells_tumor,cells_DICS)) + 
    NoAxes() + 
    ggplot2::scale_color_viridis_c(option = "plasma", oob = scales::squish,
                                   limits = c(0, 6) ) +
    AddBoundary(bon_points_tumor, color_boundary = "blue", 
                linewidth_boundary = 1) 
}

patchwork::wrap_plots(p, ncol = 1)
```

Figure 5d. Tumor centroid weights.
```{r fig5d, fig.width=4, fig.height=2.8}
cen_weights_tumor <- ComputeCentroidWeights(coords, cell_ids = cells_tumor)

PlotWeights(coords, weights = cen_weights_tumor) + NoAxes()
```

Spatial DE of tumor cells along centroid weights
```{r}
tab_cen_tumor <- RunSpatialDE(exp_mat = logNorm_expr, 
                             cell_ids = cells_tumor,
                             spatial_distance = cen_weights_tumor)
head(tab_cen_tumor[,c(1:3,7:9)])
```

Figure 5e. Top DE genes along tumor centroid weights.
```{r fig5e, fig.width=4, fig.height=2.8}
df <- tab_cen_tumor[1:10,]
df <- dplyr::arrange(df, desc(abs(Z1)))
df <- dplyr::mutate(df, gene = factor(gene, levels = gene[order(Z1)]))

ggplot(df, aes(x = gene, y = Z1, fill = trend)) +
  geom_col() +
  coord_flip() +
  scale_fill_manual(values = c("Positive" = "firebrick1", 
                               "Negative" = "cornflowerblue")) +
  labs(x = "Gene", y = "Z1", fill="Trend",
       title = "Top tumor spatial DE genes") +
  theme_minimal()
```

Spatial enrichment analysis near centroid
```{r}
sei_cen_tumor <- 
  ComputeSpatialEnrichmentIndex(exp_mat = logNorm_expr[,cells_tumor],
                                weights = cen_weights_tumor)
head(sei_cen_tumor)
```


Figure 5f. Top spatially enriched genes in tumor cells based on normalized SEI.
```{r fig5f, fig.width=4, fig.height=2.8}
df <- sei_cen_tumor[1:10,]
df <- dplyr::mutate(df, gene = factor(gene, levels = df$gene) )

ggplot(df, aes(x = gene, y = 1000*(normalized_SEI - 1) )) +
  geom_col( fill = "mediumturquoise") +
  coord_flip() +
  labs(x = "Gene", y = "1000*(Norm_SEI - 1)", 
       title = "Genes with top normalized SEI")  +
  theme_minimal()
```



Figure 5g. Expression of CEACAM6, AGR3 and DUSP2 in tumor and DCIS cells.
```{r fig5g, fig.width=6.4, fig.height=7.2}
my_genes = c("CEACAM6", "AGR3", "DUSP2")
p <- list()
for (i in my_genes) {

  p[[i]] <- 
    PlotExpression(data = coords,
                   exp_mat = logNorm_expr,
                   genes = i,
                   split_by = "cluster", 
                   point_size = 0.1,
                   sub_plot = TRUE,
                   sub_cells = c(cells_tumor,cells_DICS)) + 
    NoAxes() + 
    ggplot2::scale_color_viridis_c(option = "plasma", oob = scales::squish,
                                   limits = c(0, 6) ) +
    AddBoundary(bon_polys_dcis[1,], color_boundary = "red", 
                linewidth_boundary = 1) 
}

patchwork::wrap_plots(p, ncol = 1)
```


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Supplementary Figure 2

Figure S2a. Expression of TENT5C split by clusters.
```{r figS2a, fig.width=8, fig.height=6}
PlotExpression(data = coords, 
               exp_mat = logNorm_expr,
               genes = "TENT5C", 
               point_size = 0.1,
               split_by = "cluster") +
  NoAxes() + 
  ggplot2::scale_color_viridis_c(
    option = "plasma",
    limits = c(0, 6),
    oob = scales::squish)
```

Figure S2b. Expression of MZB1 split by clusters.
```{r figS2b, fig.width=8, fig.height=6}
PlotExpression(data = coords, 
               exp_mat = logNorm_expr,
               genes = "MZB1", 
               point_size = 0.1,
               split_by = "cluster") +
  NoAxes() + 
  ggplot2::scale_color_viridis_c(
    option = "plasma",
    limits = c(0, 6),
    oob = scales::squish)
```

<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

## Session information

The information of current R session and packages used for this analysis is shown below.

```{r}
sessionInfo()
```


\pagebreak
