---
title: "SpNeigh analysis of human healthy liver MERFISH data"
author:
  - name: Jinming Cheng
    affiliation:
    - &duk_nus Centre for Quantitative Medicine,
               Duke-NUS Medical School,
               Singapore, 169857, 
               Singapore
    # email: jinming.cheng@outlook.com
    

date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{SpNeigh analysis of mouse brain Xenium data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}

output:
  BiocStyle::html_document:
    toc_float: false
bibliography-style: plain
---


```{r,include=FALSE}
figure_path = "figures/vigette_human_liver_merfish/"
knitr::opts_chunk$set(
    eval = TRUE,
    collapse = TRUE,
    comment = "#>",
    out.width = "100%",
    dev = "png",
    dpi = 72,
    fig.height = 4.2,
    fig.width = 5.6,
    fig.path = figure_path
)
```


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


# Human healthy liver MERFISH

This Rmarkdown file includes the SpNeigh analysis of human healthy liver MERFISH dataset to reproduce all the related figures in the SpNeigh manuscript.

## Data preparation

Load packages
```{r, message=FALSE, warning=FALSE}
library(SpNeigh)
library(Seurat)
library(ggplot2)
library(dplyr)
```


Read in processed Seurat object
```{r}
seu_spatial <- readRDS("data/Human_healthy_liver_merfish.rds")
dim(seu_spatial)
```

Extract hepatocyte (Annotation from 
[Watson et al](https://datadryad.org/dataset/doi:10.5061/dryad.37pvmcvsg) )
```{r}
keep_cell_types <- c("Hepatocyte 1", "Hepatocyte 2",  "Hepatocyte 3")
cells_hepatocyte <- rownames(subset(seu_spatial@meta.data,
                                    Cell_Type %in% keep_cell_types) )
seu_spatial <- seu_spatial[,cells_hepatocyte]
dim(seu_spatial)
```

Set author-annotated cell type as cluster
```{r}
seu_spatial$seurat_clusters <- FactorNaturalOrder(seu_spatial$Cell_Type)
table(seu_spatial$seurat_clusters)
```



Extract coordinates from MERFISH seurat object
```{r}
coords <- SpNeigh::ExtractCoords(seu_spatial)
rownames(coords) <- coords$cell
head(coords)
```

Add zone information
```{r}
raw_levels <- c("Hepatocyte 1", "Hepatocyte 2", "Hepatocyte 3")
new_labels <- c("Zone 1 (Periportal)", 
                "Zone 2 (mid-lobular)", 
                "Zone 3 (Pericentral)") 
coords$zone <- factor(coords$cluster, levels = raw_levels, labels = new_labels)
table(coords$zone)
```



<!-- Define function to sort factor levels by decreasing number -->
<!-- ```{r} -->
<!-- FactorSorted = function(x, decreasing = TRUE){ -->
<!--   x = factor(x) -->
<!--   ordered_levels = sort(table(x),decreasing = decreasing) -->
<!--   x = factor(x, levels = names(ordered_levels) ) -->
<!--   x -->
<!-- } -->
<!-- ``` -->


Set colors
```{r}
my_colors <- my_colors_15[c(6,5,7)]
```

Extract normalized expression data
```{r}
logNorm_expr <- Seurat::GetAssayData(seu_spatial)
```

```{r}
class(logNorm_expr)
```


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->



<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


## SpNeigh analysis


### Figure 6

Figure 6a-1. Spatial plot.
```{r fig6a_1, fig.width=5, fig.height=3}
PlotBoundary(coords, colors = my_colors) + NoAxes()
```

Figure 6a-2. Spatial plot with boundaries.
```{r fig6a_2, fig.width=4, fig.height=3, out.width="70%"}
bon_points_1 <- GetBoundary(seu_spatial, one_cluster = "Hepatocyte 1", eps=40)
bon_points_2 <- GetBoundary(seu_spatial, one_cluster = "Hepatocyte 2", eps=90)
bon_points_3 <- GetBoundary(seu_spatial, one_cluster = "Hepatocyte 3", eps=80)

bon_polys_1 <- BuildBoundaryPoly(bon_points_1)
bon_polys_2 <- BuildBoundaryPoly(bon_points_2)
bon_polys_3 <- BuildBoundaryPoly(bon_points_3)

PlotBoundary(coords, colors = my_colors) + 
  NoAxes() + NoLegend() +
  AddBoundary(bon_points_1, color_boundary = "blue") +
  AddBoundary(bon_points_2, color_boundary = "red") +
  AddBoundary(bon_points_3, color_boundary = "yellow2")
```

Figure 6b. Spatial weights.
```{r fig6b, fig.width=12, fig.height=3}
# compute weights
weights_bon1 <- 
  ComputeBoundaryWeights(data = seu_spatial,
                         cell_ids = colnames(seu_spatial), 
                         boundary = bon_polys_1)
weights_bon2 <- 
  ComputeBoundaryWeights(data = seu_spatial,
                         cell_ids = colnames(seu_spatial), 
                         boundary = bon_polys_2)
weights_bon3 <- 
  ComputeBoundaryWeights(data = seu_spatial,
                         cell_ids = colnames(seu_spatial), 
                         boundary = bon_polys_3)
# plot weights
p <- list()
for (i in 1:3) {
  p[[i]] <- PlotWeights(seu_spatial, 
                       weights = get(paste0("weights_bon",i)),
                       point_size = 0.1) + 
    labs(title = paste0("Spatial weights ",i)) + 
    NoAxes()
}

patchwork::wrap_plots(p)
```

Spatial DE analysis
```{r}
tab_bon1 <- RunSpatialDE(exp_mat = logNorm_expr,
                         cell_ids = cells_hepatocyte,
                         spatial_distance = weights_bon1)

tab_bon2 <- RunSpatialDE(exp_mat = logNorm_expr,
                         cell_ids = cells_hepatocyte,
                         spatial_distance = weights_bon2)

tab_bon3 <- RunSpatialDE(exp_mat = logNorm_expr, 
                         cell_ids = cells_hepatocyte,
                         spatial_distance = weights_bon3)
```

Top DE genes along spatial weight 1
```{r}
head(tab_bon1)[,c(1:3,7,8)]
```

Top DE genes along spatial weight 2
```{r}
head(tab_bon2)[,c(1:3,7,8)]
```

Top DE genes along spatial weight 2
```{r}
head(tab_bon3)[,c(1:3,7,8)]
```

Figure 6c. Expression plots of selected genes.
```{r fig6c, fig.width=8, fig.height=8, warning=FALSE, message=FALSE}
my_genes <- c("ASS1", "SDS", "CYP1A2", "CYP3A4")
boundary_colors <- c("blue", "red", "yellow2")
p_genes <- list()
p <- list()
for (one_gene in my_genes) {
  for (i in 1:3) {
    p[[i]] <- PlotExpression(data = coords,
                             exp_mat = logNorm_expr, 
                             sub_plot = TRUE,
                             one_cluster = paste0("Hepatocyte ",i),
                             genes = one_gene) +
      ggplot2::scale_color_viridis_c(
        option = "plasma", limits = c(3, 6), 
        oob = scales::squish) +
      AddBoundary(get(paste0("bon_polys_",i)),
                  color_boundary = boundary_colors[i],
                  linewidth_boundary = 0.6) + 
      NoAxes() +
      theme(plot.margin = margin(0,0,0,0))
    
    # remove legend except for the last plot
    if(i!=3) {
      p[[i]] <- p[[i]] + NoLegend()
      }
  }
  p_genes[[one_gene]] <- patchwork::wrap_plots(p) + 
    patchwork::plot_layout(
      ncol = 3, 
      widths = c(3,3,4) & theme(plot.margin = margin(0,0,0,0)))
}

patchwork::wrap_plots(p_genes, ncol = 1)
```

Figure 6d-1. Scatter plots showing normalized expression of 
ASS1 versus spatial weights 1.
```{r fig6d_1, fig.width=3.5, fig.height=3}
one_gene = "ASS1"
df <- data.frame(exp = as.vector(logNorm_expr[one_gene,]),
                 weights = weights_bon1,
                 cluster = coords$cluster)

# Spearman coefficient
coef <- cor(df$exp, df$weights, method = "spearman")
coef <- round(coef, 2)

ggplot(data = df, aes(x=exp,y=weights)) + 
    geom_point(aes( color = cluster), size = 0.1) + 
    my_theme_ggplot() + 
    labs(color = "Cluster",
         title = bquote("Spearman " * rho == .(coef)),
         x = paste0("Expression of ", one_gene),
         y = "Spatial weights 1") +
  coord_cartesian(xlim=c(3.6,6.6)) +
  scale_color_manual(values = my_colors) +
  geom_smooth(method = "lm", se=FALSE) + 
  NoLegend()
```
Figure 6d-2. Scatter plots showing normalized expression of 
CYP1A2 versus spatial weights 2.
```{r fig6d_2, fig.width=3.5, fig.height=3}
one_gene = "CYP1A2"
df <- data.frame(exp = as.vector(logNorm_expr[one_gene,]),
                 weights = weights_bon2,
                 cluster = coords$cluster)

# Spearman coefficient
coef <- cor(df$exp, df$weights, method = "spearman")
coef <- round(coef, 2)

ggplot(data = df, aes(x=exp,y=weights)) + 
    geom_point(aes( color = cluster), size = 0.1) + 
    my_theme_ggplot() + 
    labs(color = "Cluster",
         title = bquote("Spearman " * rho == .(coef)),
         x = paste0("Expression of ", one_gene),
         y = "Spatial weights 2") +
  coord_cartesian(xlim=c(2,7)) +
  scale_color_manual(values = my_colors) +
  geom_smooth(method = "lm", se=FALSE) + 
  NoLegend()
```

Figure 6d-3. Scatter plots showing normalized expression of 
CYP3A4 versus spatial weights 3.
```{r fig6d_3, fig.width=3.5, fig.height=3}
one_gene = "CYP3A4"
df <- data.frame(exp = as.vector(logNorm_expr[one_gene,]),
                 weights = weights_bon3,
                 cluster = coords$cluster)

# Spearman coefficient
coef <- cor(df$exp, df$weights, method = "spearman")
coef <- round(coef, 2)

ggplot(data = df, aes(x=exp,y=weights)) + 
    geom_point(aes( color = cluster), size = 0.1) + 
    my_theme_ggplot() + 
    labs(color = "Cluster",
         title = bquote("Spearman " * rho == .(coef)),
         x = paste0("Expression of ", one_gene),
         y = "Spatial weights 3") +
  coord_cartesian(xlim=c(2,7)) +
  scale_color_manual(values = my_colors) +
  geom_smooth(method = "lm", se=FALSE) + 
  NoLegend()
```

Figure 6e. Heatmap of average gene expression across spatial weight bins.
```{r fig6e, fig.width=8, fig.height=3}
p1 <- PlotSpatialExpression(exp_mat = logNorm_expr,
                            spatial_distance = weights_bon1,
                            n_bins = 30,
                            genes = "ASS1") + 
  coord_flip() + labs(y=NULL, x="Spatial weights 1")
p2 <- PlotSpatialExpression(exp_mat = logNorm_expr,
                            spatial_distance = weights_bon2,
                            n_bins = 30,
                            genes = "CYP1A2") + 
  coord_flip() + labs(y=NULL, x="Spatial weights 2")
p3 <- PlotSpatialExpression(exp_mat = logNorm_expr,
                            spatial_distance = weights_bon3,
                            n_bins = 30,
                            genes = "CYP3A4") + 
  coord_flip() + labs(y=NULL, x="Spatial weights 3")
patchwork::wrap_plots(p1,p2,p3, ncol = 3)
```
<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->

### Supplementary Figure 3

Figure S3a-1. Spatial expression patterns of top spatially varying genes 
in liver hepatocytes (Spatial weights 1).
```{r figureS3a_1, fig.width=6, fig.height=4}
my_genes <- tab_bon1$gene[1:10]
PlotSpatialExpression(exp_mat = logNorm_expr, 
                      spatial_distance = weights_bon1,
                      genes = my_genes, 
                      scale_method = "minmax",
                      n_bins = 30,
                      n_labels = 5,
                      label_x = "Spatial weights 1") 
```

Figure S3a-2. Spatial expression patterns of top spatially varying genes 
in liver hepatocytes (Spatial weights 2).
```{r figureS3a_2, fig.width=6, fig.height=4}
my_genes <- tab_bon2$gene[1:10]
PlotSpatialExpression(exp_mat = logNorm_expr, 
                      spatial_distance = weights_bon2,
                      genes = my_genes, 
                      scale_method = "minmax",
                      n_bins = 30,
                      n_labels = 5,
                      label_x = "Spatial weights 2") 
```


Figure S3a-3. Spatial expression patterns of top spatially varying genes 
in liver hepatocytes (Spatial weights 3).
```{r figureS3a_3, fig.width=6, fig.height=4}
my_genes <- tab_bon3$gene[1:10]
PlotSpatialExpression(exp_mat = logNorm_expr, 
                      spatial_distance = weights_bon3,
                      genes = my_genes, 
                      scale_method = "minmax",
                      n_bins = 30,
                      n_labels = 5,
                      label_x = "Spatial weights 3") 
```

Figure S3b. Spatial expression plots of GLS2, CYP2E1, and COL1A1.
```{r figS3bc, fig.width=8, fig.height=6, warning=FALSE, message=FALSE}
my_genes <- c("GLS2", "CYP2E1", "COL1A1")
boundary_colors <- c("blue", "red", "yellow2")
p_genes <- list()
p <- list()
for (one_gene in my_genes) {
  for (i in 1:3) {
    p[[i]] <- PlotExpression(data = coords,
                             exp_mat = logNorm_expr, 
                             sub_plot = TRUE,
                             one_cluster = paste0("Hepatocyte ",i),
                             genes = one_gene) +
      ggplot2::scale_color_viridis_c(
        option = "plasma", limits = c(0, 6), 
        oob = scales::squish) +
      AddBoundary(get(paste0("bon_polys_",i)),
                  color_boundary = boundary_colors[i],
                  linewidth_boundary = 0.6) + 
      NoAxes() +
      theme(plot.margin = margin(0,0,0,0))
    
    # remove legend except for the last plot
    if(i!=3) {
      p[[i]] <- p[[i]] + NoLegend()
      }
  }
  p_genes[[one_gene]] <- patchwork::wrap_plots(p) + 
    patchwork::plot_layout(
      ncol = 3, 
      widths = c(3,3,4) & theme(plot.margin = margin(0,0,0,0)))
}

patchwork::wrap_plots(p_genes, ncol = 1)
```
<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->


<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->
<!-- %                                                 % -->


<!-- %                                                 % -->
<!-- %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% -->

## Session information

The information of current R session and packages used for this analysis is shown below.

```{r}
sessionInfo()
```


\pagebreak
